Fail Faster

### 2025-03-06 paper/paper.tex
push
more edits to structure

@@

\subsection{Property-Based Testing}
\jwc{Mention the shape of PBT dsls, give running example here (or maybe earlier!). Monadic dsls.}

\begin{lstlisting}
let int_pair : (int * int) Bq.t =
  let%bind x = int 0 100 in
  let%bind y = int 0 x in
  return (x,y)
\end{lstlisting}

\jwc{\texttt{int} here is really \texttt{int\_uniform\_inclusive}. I also think we should completely omit named arguments in the paper so we don't confuse with splices.}

@@

\subsection{Other Languages and Libraries}

@@

\jwc{
  \begin{itemize}
    \item How does BQ work?
    \begin{itemize}
      \item The basic generator type: \texttt{'a generator = int -> SR.t -> 'a}. Size and random seed to deterministic value. (note: \texttt{SR.t} is a mutable seed)
      \item Also two inverse functions, \texttt{create}, and \texttt{generate}.
      \item This gets a monad intance in the obvious way (show code).
      \item Also show the code for \texttt{int}, how it calls the underlying SR function.
    \end{itemize}
    \item Note that \emph{extensionally} \texttt{generate (create (fun size random -> e)) size random = e}, but the OCaml compiler does not always perform this optimization, or do the inlining required to expose it.
    (When sufficiently obfuscated behind returns and binds ...) This program compiles to code that (1) allocates the closure for `e', (2) passes it to create (which returns the closure), and then calls (3) generate, which immediately jumps into the closure.
    \item Show allocation benchmarks of the running example, versus the version where you inline everything.
  \end{itemize}
}

@@

\jwc{
  \begin{itemize}
    \item To eliminate this runtime overhead and generate efficient code, we do a usual ``binding-time analysis''.
    \item The crux of this binding-time analysis is that while generators are `just' values of type \texttt{'a t}, in practice they are always \emph{completely statically defined}: one never
    computes a generator dynamically at test time.
  \end{itemize}
}

@@

@@

\begin{lstlisting}
type 'a gen = int code -> Random.t code -> 'a code

let return (cx : 'a code) : 'a gen = fun size_c random_c -> cx

let bind (g : 'a gen) (k : 'a code -> 'b gen) : 'b gen =
  fun size_c random_c ->
    .<
      let x = .~(g size_c random_c) in
      .~(
        (k .<x>.) size_c random_c
      )
    >.

let int (lo : int code) (hi : int code) : int code gen =
  fun size_c random_c ->
    .< SR.int .~random_c .~lo .~hi >.
\end{lstlisting}

@@

\begin{lstlisting}
let to_bq (g : 'a code gen) : ('a Bq.t) code =
  .<
    Bq.create (fun size random ->
      .~(g .<size>. .<random>.)
    )
  >.
\end{lstlisting}

@@

\begin{lstlisting}
let int_pair_staged : ((int * int) code) gen =
  bind (int .<0>. .<100>.) (fun cx ->
    bind (int .<0> cx) (fun cy ->
      return .<(.~cx,.~cy)>.
    )
  )
let int_pair : (int * int) Bq.t code = to_bq int_pair_staged
====
.<
  Bq.create (fun size random ->
    let x = SR.int random 0 100 in
    let y = SR.int random 0 x in
    (x,y)
  )
.>
\end{lstlisting}

\jwc{So if you write with the \texttt{gen} combinators and then only call \texttt{to\_bq} at the end, you get something that includes no creates/generates!}

@@

>>>>

\lstset{
  basicstyle=\small\fontfamily{cmtt}\selectfont, % Computer Modern Typewriter
}

@@

\jwc{In this paper, we'll use OCaml to illustrate the ideas, but the concepts are portable to other languages: see Section~\ref{subsection:other-langs}}

\subsection{Property-Based Testing}
\jwc{Usual spiel about pbt and generators: here's the BQ generator library, it has return and bind, this is what they're for.}

\begin{lstlisting}
module Bq = struct
  type 'a t = int -> SR.t -> 'a

  let return (x : 'a) : 'a t = fun size seed -> x

  let bind (g : 'a t) (f : 'a -> 'b t) : 'b t =
    fun size seed ->
      let x = g size seed in
      (f x) size seed

  let gen_int (lo : int) (hi : int) : int t =
    fun size seed -> SR.int seed lo hi
end
\end{lstlisting}

\jwc{\texttt{gen\_int} here is really \texttt{int\_uniform\_inclusive}. I also think we should completely omit named arguments in the paper so we don't confuse with splices.}

\jwc{Here's an example of a generator: you can sample from it and it gives you pairs of ints, one less than the other.}

\begin{lstlisting}
let int_pair : (int * int) Bq.t =
  let%bind x = Bq.gen_int 0 100 in
  let%bind y = Bq.gen_int 0 x in
  return (x,y)
\end{lstlisting}

\jwc{This desugars to...}

\begin{lstlisting}
let int_pair : (int * int) Bq.t =
  Bq.bind (Bq.gen_int 0 100) (fun x ->
    Bq.bind (Bq.gen_int 0 x) (fun y ->
      Bq.return (x,y)
  ))
\end{lstlisting}

@@

\jwc{Staged metaprogramming has a long history as a technique for eliminating the runtime overhead of abstractions,
specializing programs at an earlier stage to run more efficiently at a later stage.}

\jwc{
  \begin{itemize}
    \item The \texttt{'a code} type, quote, splice, stage distinction.
    \item The difference between \texttt{('a -> 'b) code} and \texttt{'a code -> 'b code}
  \end{itemize}
}

\subsection{Other Languages and Libraries}
\label{subsection:other-langs}

@@

\jwc{
  \begin{itemize}
    \item How does BQ work?
    \begin{itemize}
      \item The basic generator type: \texttt{'a generator = int -> SR.t -> 'a}. Size and random seed to deterministic value. (note: \texttt{SR.t} is a mutable seed)
      \item Also two inverse functions, \texttt{create}, and \texttt{generate}.
      \item This gets a monad intance in the obvious way (show code).
      \item Also show the code for \texttt{int}, how it calls the underlying SR function.
    \end{itemize}
    \item Note that \emph{extensionally} \texttt{generate (create (fun size random -> e)) size random = e}, but the OCaml compiler does not always perform this optimization, or do the inlining required to expose it.
    (When sufficiently obfuscated behind returns and binds ...) This program compiles to code that (1) allocates the closure for `e', (2) passes it to create (which returns the closure), and then calls (3) generate, which immediately jumps into the closure.
    \item Show allocation benchmarks of the running example, versus the version where you inline everything.
    \item Let's look at the running example: inline it all the way.
  \end{itemize}
}

\begin{lstlisting}
let int_pair : (int * int) Bq.t =
  Bq.bind (Bq.gen_int 0 100) (fun x ->
    Bq.bind (Bq.gen_int 0 x) (fun y ->
      Bq.return (x,y)
  ))
\end{lstlisting}

@@

\jwc{Usual introduction to this section, corresponding to how we talked about it in the intro}

\jwc{
  \begin{itemize}
    \item To eliminate this runtime overhead and generate efficient code, we do a usual ``binding-time analysis''.
    \item The crux of this binding-time analysis is that while generators are `just' values of type \texttt{'a t}, in practice they are always \emph{completely statically defined}: one never
    computes a generator dynamically at test time.
    \item The size and the random seed are only known at run-time (the later stage), but the code of the generator itself is known at compile time.
    \item This means the generator type should be \texttt{int code -> SR.t code -> 'a code}: a statically known-function from dynamically-known int and seed to dynamically-determined result.
  \end{itemize}
}

@@

A generator of \texttt{'a}s is then a (compile-time) function \texttt{int code -> Random.t code -> 'a code}.

@@

\jwc{
  \begin{itemize}
    \item The simplest combinators are presented below.
    \item \texttt{return} is the constant generator, given code for \texttt{x : 'a}, it gives the generator which always generates \texttt{x}
    \item \texttt{bind} is like BQ's bind --- sequencing generators by passing the result of one to another --- except that the continuation gets access to the
    \texttt{code} for the result of running the first generator. At compile time, we know only that \texttt{g} will return \emph{some} \texttt{a}, but we do not get access to the particular \texttt{a}.
  \end{itemize}
}

\begin{lstlisting}

module Gen = struct
  type 'a t = int code -> Random.t code -> 'a code

  let return (cx : 'a code) : 'a t = fun size_c random_c -> cx

  let bind (g : 'a t) (k : 'a code -> 'b t) : 'b t =
    fun size_c random_c ->
      .<
        let x = .~(g size_c random_c) in
        .~(
          (k .<x>.) size_c random_c
        )
      >.

  let int (lo : int code) (hi : int code) : int t =
    fun size_c random_c ->
      .< SR.int .~random_c .~lo .~hi >.
end
\end{lstlisting}

@@

\begin{lstlisting}
let to_bq (g : 'a code Gen.t) : ('a Bq.t) code =
  .<
    Bq.create (fun size random ->
      .~(g .<size>. .<random>.)
    )
  >.
\end{lstlisting}

@@

\begin{lstlisting}
let int_pair_staged : ((int * int) code) Gen.t =
  Gen.bind (int .<0>. .<100>.) (fun cx ->
    Gen.bind (int .<0> cx) (fun cy ->
      Gen.return .<(.~cx,.~cy)>.
    )
  )
let int_pair : (int * int) Bq.t code = Gen.to_bq int_pair_staged
====
.<
  Bq.create (fun size random ->
    let x = SR.int random 0 100 in
    let y = SR.int random 0 x in
    (x,y)
  )
.>
\end{lstlisting}

\jwc{So if you write with the \texttt{Gen} combinators and then only call \texttt{to\_bq} at the end, you get something that includes no creates/generates!}

@@

\jwc{This is 3-stage metaprogramming! There's PPX-time, compile time, and run-time.}
